The Changing Shape of Empire: A GM Study of Chimú Bottles in Museum Collections

### 2022-02-04 supp/04-gm.neck.Rmd
<init neck.morph>

# Bottles vs Jar morphology

@@

```{r load.data, echo = TRUE}
# load packages
#devtools::install_github("mlcollyer/RRPP")
#devtools::install_github("geomorphR/geomorph", ref = "Stable", build_vignettes = TRUE)
library(here)
library(geomorph)
library(tidyverse)
library(wesanderson)

@@

```{r gpa, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# gpa
Y.gpa <- gpagen(coords, 
                PrinAxes = TRUE, 
                ProcD = TRUE, 
                Proj = TRUE, 
                print.progress = FALSE)

@@

```{r box1, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE, fig.cap = "Boxplot of centroid size by spatial/temporal unit."}
# attributes for boxplot
csz <- qdata$csz
group <-  qdata$group
type <- qdata$type

# boxplot of vessel centroid size by type
csz.type <- ggplot(qdata, aes(x = type, y = csz, color = type)) + 
  geom_boxplot(notch = FALSE) +
  geom_dotplot(binaxis = 'y', stackdir = 'center', dotsize = 0.3) +
  scale_colour_manual(values = wes_palette("Moonrise2")) +
  theme(legend.position = "none") +
  labs(x = 'Group', y = 'Centroid Size')

@@

Principal components analysis [@RN1746] was used to visualise shape variation among the bottles. The shape changes described by each principal axis are commonly visualised using thin-plate spline warping of a reference 3D mesh [@RN1731;@RN479].

```{r gmpca, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# principal components analysis
pca<-gm.prcomp(Y.gpa$coords)
summary(pca)

@@

```{r pca.warp, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE, fig.cap="Results of PCA summarising shape variation in the aggregated sample; gray, north-formative/early; orange, north-late/historic; brown, south-formative/early; and black, south-late/historic; with mean shapes (gray) contrasted with max/min shapes (black)."}
# pca warp                   
#knitr::include_graphics('images/pca-warp.jpg')
```

## Vessel allometry

```{r allometry.1, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# allometry: shape as a function of vessel size?
fit.size<-procD.lm(shape ~ size, 
                   data = gdf, 
                   print.progress = FALSE, 
                   iter = 9999)

# Procrustes ANOVA: does shape change with size?
anova(fit.size)
```

```{r common.allometry.1, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# common allometries
fit.common<-procD.lm(shape ~ size + type, 
                     data = gdf, 
                     print.progress = FALSE, 
                     iter = 9999)

anova(fit.common)

common.allometry <- plot(fit.size,
                         type = "regression",
                         reg.type = "RegScore",
                         predictor = log(gdf$size),
                         pch = pch.gps.type,
                         col = col.gps.type)

# common allometric component (Mitteroecker 2004)
plotAllometry(fit.size, 
              size = gdf$size, 
              logsz = TRUE, 
              method = "CAC", 
              pch = pch.gps.type, 
              col = col.gps.type)
```

```{r unique.allometry.1, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# unique allometries
fit.unique<-procD.lm(shape ~ size * type, 
                     data = gdf, 
                     print.progress = FALSE, 
                     iter = 9999)

anova(fit.unique)

# do vessel forms express parallel, convergent, or divergent morphological characteristics?
plotAllometry(fit.unique, 
              size = gdf$size, 
              logsz = TRUE, 
              method = "PredLine", 
              pch = pch.gps.type, 
              col = col.gps.type)
```

## Vessel size and shape

A residual randomisation permutation procedure (RRPP; n = 10,000 permutations) was used for all Procrustes ANOVAs [@RN1655;@RN11775], which has higher statistical power and a greater ability to identify patterns in the data should they be present [@RN1719]. To assess whether shape changes differ by group (geography and time), Procrustes ANOVAs [@RN1749] were also run that enlist effect-sizes (zscores) computed as standard deviates of the generated sampling distributions [@RN1756].

```{r def.mod.1, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# size as a function of vessel form (bottle/jar)?
fit.size.type <- procD.lm(size ~ type,
                data = gdf,
                print.progress = FALSE,
                iter = 9999)

# Procrustes ANOVA: do sizes differ as a function of vessel form?
anova(fit.size.type)

# shape as a function of vessel form?
fit.shape.type <- procD.lm(shape ~ type,
                data = gdf,
                print.progress = FALSE,
                iter = 9999)

# Procrustes ANOVA: do shapes differ as a function of vessel form?
anova(fit.shape.type)
```

## Morphological disparity

```{r disparity, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# morphological disparity: do either of the groups display greater 
# size variation among individuals relative to the other group?
morphol.disparity(fit.size.type, 
                  groups = qdata$type, 
                  data = gdf, 
                  print.progress = FALSE, 
                  iter = 9999)

# morphological disparity: do either of the groups display greater 
#shape variation among individuals relative to the other group?
morphol.disparity(fit.shape.type, 
                  groups = qdata$type, 
                  data = gdf, 
                  print.progress = FALSE, 
                  iter = 9999)
```

@@

```{r mshape.h1, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# subset landmark coordinates to produce mean shapes
new.coords<-coords.subset(A = Y.gpa$coords,
                          group = qdata$type)
names(new.coords)

# group shape means
mean <- lapply(new.coords, mshape)

@@

knitr::include_graphics('images/mshape.png')
```

>>>>

# 3DGM: Neck morphology

@@

```{r load.data.s, echo = TRUE}
# load packages
#devtools::install_github("mlcollyer/RRPP")
#devtools::install_github("geomorphR/geomorph", ref = "Stable", build_vignettes = TRUE)
library(here)
library(geomorph)
library(tidyverse)
library(wesanderson)

@@

```{r gpa.s, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
#select landmarks associated with neck
a <- c(1:10, 36:45)

# new coords neck
neck.coords <- coords[a,,]

# gpa
Y.gpa <- gpagen(neck.coords, 
                PrinAxes = TRUE, 
                ProcD = TRUE, 
                Proj = TRUE, 
                print.progress = FALSE)

@@

```{r box1.s, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE, fig.cap = "Boxplot of centroid size by spatial/temporal unit."}
# attributes for boxplot
csz <- qdata$csz
type <-  qdata$type

# palette
pal <- wes_palette("Moonrise2", 2, type = "continuous")

# boxplot of vessel centroid size by type
csz.type <- ggplot(qdata, aes(x = type, y = csz, color = type)) + 
  geom_boxplot(notch = FALSE) +
  geom_dotplot(binaxis = 'y', stackdir = 'center', dotsize = 0.3) +
  scale_colour_manual(values = pal) +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(x = 'Type (Neck)', 
       y = 'Centroid Size')

@@

Principal components analysis [@RN1746] was used to visualise shape variation among bottle and jar necks. The shape changes described by each principal axis are commonly visualised using thin-plate spline warping of a reference 3D mesh [@RN1731;@RN479].

```{r gmpca.s, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# principal components analysis
pca<-gm.prcomp(Y.gpa$coords)
summary(pca)

@@

```{r pca.warp.s, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE, fig.cap="Results of PCA summarising shape variation in the aggregated sample; gray, north-formative/early; orange, north-late/historic; brown, south-formative/early; and black, south-late/historic; with mean shapes (gray) contrasted with max/min shapes (black)."}
# pca warp                   
#knitr::include_graphics('images/pca-warp.jpg')
```

## Neck size and shape

A residual randomisation permutation procedure (RRPP; n = 10,000 permutations) was used for all Procrustes ANOVAs [@RN1655;@RN11775], which has higher statistical power and a greater ability to identify patterns in the data should they be present [@RN1719]. To assess whether shape changes differ by type (geography and time), Procrustes ANOVAs [@RN1749] were also run that enlist effect-sizes (zscores) computed as standard deviates of the generated sampling distributions [@RN1756].

```{r def.mod.1.s, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# neck size as a function of vessel form (bottle/jar)?
fit.size.type <- procD.lm(size ~ type,
                            data = gdf,
                            print.progress = FALSE,
                            iter = 9999)

# Procrustes ANOVA: do neck sizes differ as a function of vessel form?
anova(fit.size.type)

# neck shape as a function of vessel form?
fit.shape.type <- procD.lm(shape ~ type,
                data = gdf,
                print.progress = FALSE,
                iter = 9999)

# Procrustes ANOVA: do neck shapes differ as a function of vessel form?
anova(fit.shape.type)
```

@@

```{r mshape.s, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# subset landmark coordinates to produce mean shapes
new.coords<-coords.subset(A = Y.gpa$coords,
                          group = qdata$type)
names(new.coords)

# type shape means
mean <- lapply(new.coords, mshape)

@@

knitr::include_graphics('images/mshape-neck.png')
```

### 2022-02-04 supp/index.Rmd
<init neck.morph>

### Geometric morphometrics

 * Bottles and jars **differ significantly** in shape
 * Bottles and jars do not differ in size

@@

>>>>

### 3D Geometric morphometrics

**Whole vessel morphology**:
 * Bottles and jars **differ significantly** in shape
 * Bottles and jars do not differ in size

@@

**Neck morphology**:
 * Bottle and jar necks **differ significantly** in shape
 * Bottle and jar necks do not differ in size

```{r botjarneck, out.width = "100%", dpi = 300, echo=FALSE, warning=FALSE, fig.cap="Mean shapes for bottle necks (a), jar necks (b), and comparison of bottle (gray) and jar (black) neck shapes (c)."}
knitr::include_graphics('images/mshape-neck.png')
```
